---
title: "Simulation Based on Real-World Data"
author: "Lujun Zhang and Jun Chen"
date: "2024-08-14"
output:
  html_document:
    number_sections: true
---
    
# Setup

-  Required packages: 
    +  `doParallel`, `foreach`, `doRNG`, `Matrix`, `matrixStats`, `tidyverse`, `Hmisc`, `knitr`, `kableExtra` from CRAN; 
    +  `DESeq2` from bioconductor; 
    +  `ideas` from https://github.com/Sun-lab/ideas


```{r setup, include=TRUE, warning=FALSE, message=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(knitr)
library(kableExtra)
library(tidyverse)
library(Matrix)
```

# Data Inputs and Preprocessing for Simulation

The simulations are based on cell type "L2/3" found in the region of the prefrontal cortex (PFC) from an autism study conducted by Velmeshev, et al. (2019). The data have been preprocessed in accordance with the procedures outlined in the IDEAS paper and were directly obtained from the link [https://github.com/Sun-lab/ideas_pipeline/tree/main/Autism/data/ct_mtx](https://github.com/Sun-lab/ideas_pipeline/tree/main/Autism/data/ct_mtx) (Zhang et al., 2022).

In the following code chunk, we will read in the `count_matrix`, which comprises count data from single-cell RNA sequencing, as well as metadata indicating cell assignments to specific individuals and diagnoses for autism spectrum disorder (ASD) and healthy controls. We will also perform sparsity filtering, excluding genes appearing in fewer than 20% of cells with a defined threshold of `sparsity_threshold` set at 0.8. We will discard cells with insufficient sequencing depth, defined by the `thre_quant_per_cell` threshold set at 0.05 (representing the bottom 5%, equivalent to a sequencing depth of approximately 4100), and cells overly deeply sequenced, defined by the `max_quant_per_cell` threshold set at 0.99 (representing the top 1%, equivalent to a sequencing depth of approximately 41000).

The subsequent code chunk will save required datasets as "setup0702.RData" in the "real_data_sim/" folder. This image includes the processed sequencing data, metadata, and functions from the "source_code/DiSC.R" script.


```{r input}
###########################################################
#remember to copy these after modifying the count matrix
# ngene = nrow(count_matrix)
# ncell = ncol(count_matrix)
# read_depth <- colSums(count_matrix)
# cat("ngene: ", ngene, "ncell: ", ncell, "read_depth[1:10]: ", read_depth[1:10], "\n")
###########################################################
sparsity_threshold = 0.8 #or NULL. Keep genes with sparsity < SPARSITY_THRESHOLD
thre_quant_per_cell = 0.05 # 5% quantile. Or 0 to keep all the cells
max_quant_per_cell  = 0.99 # 99% quantile. Or 1 to keep all the cells
boot = 1000

fp = "ideas_pipeline-main/Autism/data/"

ctp  = "L2_3"
cat("Using cell type:", ctp, "\n")
## count_matrix, genes in rows and cells in columns
count_matrix = readRDS(file.path(sprintf("%sct_mtx/PFC_%s.rds", fp, ctp)))
ngene = nrow(count_matrix)
ncell = ncol(count_matrix)
read_depth <- colSums(count_matrix)
if(ngene < ncell) 
  warning("the number of genes is smaller than the number of cells, please check if genes are in rows")
cat("ngene: ", ngene, "ncell: ", ncell, "read_depth[1:10]: ", read_depth[1:10], "\n")
#### Sparsity filtering
if(!is.null(sparsity_threshold)){
  cat("sparsity filtering with threshold ", sparsity_threshold, " :\n")
  gene2keep = which(rowSums(count_matrix == 0) < sparsity_threshold*ncell)
  count_matrix = count_matrix[gene2keep,]
  read_depth = colSums(count_matrix)
  ngene = nrow(count_matrix)
  ncell = ncol(count_matrix)
  read_depth <- colSums(count_matrix)
  cat("ngene: ", ngene, "ncell: ", ncell, "read_depth[1:10]: ", read_depth[1:10], "\n")
}
#### discard insufficiently and overly deeply sequenced cells
if(TRUE){
  min_count = round(quantile(read_depth, probs = thre_quant_per_cell))
  max_count = round(quantile(read_depth, probs = max_quant_per_cell))
  cat(str_glue("read depth filtering with threshold (min = {thre_quant_per_cell}, max = {max_quant_per_cell}),\nwhich corresponds to read depth of (min = {min_count}, max = {max_count})"))
  count_matrix <- count_matrix[, which(read_depth >= min_count & 
                                         read_depth <= max_count)]
  ngene = nrow(count_matrix)
  ncell = ncol(count_matrix)
  read_depth <- colSums(count_matrix)
  cat("ngene: ", ngene, "ncell: ", ncell, "read_depth[1:10]: ", read_depth[1:10], "\n")
} 

## meta_cell
meta_raw = read_tsv(file.path(fp, "meta.tsv"), col_types = "cffcfnffffnnnndd")
meta_cell <- meta_raw %>% filter(region == "PFC") %>%
  filter(cluster == str_replace(ctp, "_", "/")) %>%
  dplyr::rename(`cell_id` = "cell")
if(sum(!(colnames(count_matrix) %in% meta_cell$cell_id)) >0)
  error("meta_cell and count_matrix dont match") else{
    meta_cell <- meta_cell %>% filter(cell_id %in% colnames(count_matrix))
    if(sum(meta_cell$cell_id != colnames(count_matrix)) >0 )
      error("meta_cell and count_matrix dont match2")
    meta_cell$read_depth <- read_depth
  }

## meta_ind
meta_ind <- meta_cell %>% distinct(individual, .keep_all = TRUE) %>%
  dplyr::select(-`cell_id`, -`genes`, -`UMIs`, -`RNA mitochondr. percent`,
         -`RNA ribosomal percent`, -`read_depth`) %>%
  dplyr::rename(RIN = "RNA Integrity Number")
nindi = nrow(meta_ind)

## more prepossessing
set.seed(seed = 123456)
meta_cell = as.data.frame(meta_cell) 
meta_ind = as.data.frame(meta_ind)
count_matrix = as.matrix(count_matrix)
perm.index = sapply(1:boot, simplify = TRUE,
                    FUN = function(xxxx) 
                      return(sample(1:nindi)) )

count_matrix_bulk = matrix(NA, nrow = ngene, ncol = nindi)
rownames(count_matrix_bulk) = rownames(count_matrix)
colnames(count_matrix_bulk) = meta_ind$individual
if(sum(meta_cell$cell_id != colnames(count_matrix))==0)
  for (i_ind in 1:length(meta_ind$individual)) {
    cur_ind   = meta_ind$individual[i_ind]
    cur_ind_m = count_matrix[, meta_cell$individual == cur_ind] 
    count_matrix_bulk[, i_ind] = rowSums(cur_ind_m, na.rm = TRUE)
  }else
    error("meta_cell$cell_id and colnames(count_matrix) dont match")

## save required dataset
save(list = c("meta_cell", "meta_ind", "count_matrix", "perm.index",
              "count_matrix_bulk"),
     file = "real_data_sim/setup0814.RData", compress = "xz")
```

# Simulation Study (False Discovery Rate and Type I Error)

In the following code chunk, we will generate sbatch shell scripts (.sh) and R scripts (.R) within the "real_data_sim/" directory. These files are designed for execution on a high-performance computing cluster utilizing the Slurm workload manager.

This code segment will permute labels for the variable `diagnosis` (ASD vs. control) base on `perm.index` and apply different statistical methods to investigate the association between `diagnosis` and single-cell RNA sequencing data. We aim to examine the type I error rates and false discovery rates of these methods. The results will be saved in "real_data_sim/type1_{METHOD}_{DATE}.RData."

```{r simulation FDR setup}
file.copy(from = "source_code/DiSC.R", to = "real_data_sim/DiSC.R",
          copy.date = TRUE, overwrite  = TRUE)

WORKINGDIRECTORY = "/home/guanwh/zhan7474/mayo/real_data_sim/"
boot
BOOT = as.character(boot)
DATE = "0814"
CLASS = "type1"

METHOD_S = c("disc", "ideas", "deseq")
for(METHOD in METHOD_S){
  MEM = ifelse(METHOD == "disc", 2500*2, 2500) %>% as.character()
  NCORES = min(80, round(200000/as.numeric(MEM), digits = -1)) %>% as.character()
  # shell script
  read_file("source_code/real_sim_type1.sh") %>% 
    str_replace_all(pattern = "DATE", DATE) %>%
    str_replace_all(pattern = "WORKINGDIRECTORY", WORKINGDIRECTORY) %>%
    str_replace_all(pattern = "CLASS", CLASS) %>%
    str_replace_all(pattern = "MEM", MEM) %>%
    str_replace_all(pattern = "NCORES", NCORES) %>%
    str_replace_all(pattern = "METHOD", METHOD) %>%
    write_file(file = str_glue("real_data_sim/{CLASS}_{METHOD}_{DATE}.sh"), 
               append = FALSE)
  # R script
  read_file(str_glue("source_code/real_sim_type1_{METHOD}.R")) %>% 
    str_replace_all(pattern = "DATE", DATE) %>%
    str_replace_all(pattern = "WORKINGDIRECTORY", WORKINGDIRECTORY) %>%
    str_replace_all(pattern = "BOOT", BOOT) %>%
    str_replace_all(pattern = "NCORES", NCORES) %>%
    str_replace_all(pattern = "CLASS", CLASS) %>%
    write_file(file = str_glue("real_data_sim/{CLASS}_{METHOD}_{DATE}.R"), 
               append = FALSE)
}
```

-  Example shell script (`r str_glue("real_data_sim/{CLASS}_disc_{DATE}.sh")`)

```{bash, file = str_glue("real_data_sim/{CLASS}_disc_{DATE}.sh"), eval = FALSE}
```

-  R script for DiSC (`r str_glue("real_data_sim/{CLASS}_disc_{DATE}.R")`)

```{r, file = str_glue("real_data_sim/{CLASS}_disc_{DATE}.R"), eval = FALSE}
```

-  R script for IDEAS (`r str_glue("real_data_sim/{CLASS}_ideas_{DATE}.R")`)

```{r, file = str_glue("real_data_sim/{CLASS}_ideas_{DATE}.R"), eval = FALSE}
```

-  R script for DESeq2 (`r str_glue("real_data_sim/{CLASS}_deseq_{DATE}.R")`)

```{r, file = str_glue("real_data_sim/{CLASS}_deseq_{DATE}.R"), eval = FALSE}
```

-  Run the above scripts on a high performance computer.

# Results and Tables (False discovery rate and Type I error)

-  Extracting results

```{r simulation result description}
METHOD_S = c("DiSC" = "disc", "IDEAS" = "ideas", "DESeq2" = "deseq")
meths = names(METHOD_S) #displayed names of methods
DATE
CLASS
boot

summarize_binconf <- function(x, repl){
  # calculate the binomial confidence interval with x successes and repl trials
  # returns a string
  CI_95 <- Hmisc::binconf(x = x, n = repl, method=c("wilson"))
  CI_95 <- round(CI_95*100, digits = 1)
  CI_95 <- paste(CI_95[,"PointEst"]," (", CI_95[,"Lower"], 
                 ", ", CI_95[,"Upper"],")", sep="")
  return(CI_95)
}

for(METHOD in METHOD_S){
  filename <- str_glue("real_data_sim/{CLASS}_{METHOD}_{DATE}.RData")
  if(!file.exists(filename))
    next
  load(filename) # "disc_pval", "ideas_pval", "deseq_pval"
  cat(str_glue("loaded {filename}"),"\n")
}
#create a list of "_pval" in the order of "METHOD_S" 
eval(parse(text= str_c("assign('pval_list', list(", 
                       str_c(str_c(METHOD_S, "_pval"), 
                             collapse = ","), "))"))) 
names(pval_list) <- names(METHOD_S) # displayed names of METHOD_S

pval_list <- lapply(pval_list, function(x_pval){
  #convert the disc_pval to an array
  # temp <- array(unlist(disc_pval), 
  #               dim = c(dim(disc_pval[[1]]), boot),
  #               dimnames = append(dimnames(disc_pval[[1]]), list(NULL)))
  # identical(temp[,,1], disc_pval[[1]]) #TRUE
  # identical(temp[,,3], disc_pval[[3]]) #TRUE
  return(
    array(unlist(x_pval), 
          dim = c(dim(x_pval[[1]]), boot), 
          dimnames = list(NULL, c("raw", "adjusted"), NULL))
  )# genes, (p.raw, p.adj), boot
})

nNAs_results <- as.data.frame(lapply(pval_list, FUN = function(xpval){
  colSums(is.na(xpval[,"adjusted",]))
}))
summary(nNAs_results)
```

-  summarizing tables

```{r simulation FDR Type I}
thre_raw = 0.05
thre_adj = 0.10
FDR_results <- 
  data.frame(
    method = meths,
    p_raw_mean = rep(NA_real_, length(meths)),
    p_raw_sd = rep(NA_real_, length(meths)),
    p_raw = rep(NA_character_, length(meths)),
    fdr = rep(NA_character_, length(meths)),
    ANFD_mean = rep(NA_real_, length(meths)),
    ANFD_sd = rep(NA_real_, length(meths)),
    ANFD = rep(NA_character_, length(meths))
  )
FDR_results$fdr <- pval_list %>% 
  lapply(X = ., FUN = function(x_pval)
    sum(colSums(x_pval[, "adjusted", ] <= thre_adj, na.rm = TRUE) >0)) %>%
        #at least one discovery#####################################
  unlist(.) %>%
  summarize_binconf(x = ., repl = boot)

FDR_results$ANFD_mean <- pval_list %>% 
  lapply(X = ., FUN = function(x_pval)
    mean(colSums(x_pval[, "adjusted", ] <= thre_adj, na.rm = TRUE))) %>%
         # discoveries in each bootstrap#########################
  unlist(.) 
FDR_results$ANFD_sd <- pval_list %>% 
  lapply(X = ., FUN = function(x_pval)
    sd(colSums(x_pval[, "adjusted", ] <= thre_adj, na.rm = TRUE))) %>%
       # discoveries in each bootstrap#########################
  unlist(.) 
FDR_results$ANFD <- str_glue("{round(FDR_results$ANFD_mean,1)} ({round(FDR_results$ANFD_mean-1.96*FDR_results$ANFD_sd/sqrt(boot),1)}, {round(FDR_results$ANFD_mean+1.96*FDR_results$ANFD_sd/sqrt(boot),1)})")

FDR_results$p_raw_mean <- pval_list %>%
  lapply(X = ., FUN = function(x_pval)
    mean(colMeans(x_pval[, "raw", ] <= thre_raw, na.rm = TRUE))) %>%
  unlist(.)
FDR_results$p_raw_sd <- pval_list %>%
  lapply(X = ., FUN = function(x_pval)
    sd(colMeans(x_pval[, "raw", ] <= thre_raw, na.rm = TRUE))) %>%
  unlist(.)
FDR_results$p_raw <- str_glue("{round(FDR_results$p_raw_mean*100,2)} ({round((FDR_results$p_raw_mean-1.96*FDR_results$p_raw_sd/sqrt(boot))*100,2)}, {round((FDR_results$p_raw_mean+1.96*FDR_results$p_raw_sd/sqrt(boot))*100,2)})")

# Tables
tbl.temp <- FDR_results %>%
  select(method, p_raw, fdr, ANFD) %>%
  rename("Type I error rate (%)" = p_raw) %>%
  rename("False discovery rate (%)" = fdr) %>%
  rename("Average number of false positive fundings" = ANFD) %>%
  kbl() %>%
  kable_classic(full_width = F, html_font = "Cambria")
tbl.temp
save_kable(tbl.temp, file = "results/real_null_fdr.jpeg", 
           density = 1000, zoom = 2.5)
```

# Session Info

```{r session info}
sessionInfo()
```


# References

-    Zhang, M., et al. IDEAS: individual level differential expression analysis for single-cell RNA-seq data. Genome Biol 23, 33 (2022). https://doi.org/10.1186/s13059-022-02605-1.
-    Velmeshev, D., et al. Single-cell genomics identifies cell type–specific molecular changes in autism. Science 364.6441 (2019): 685-689.
